---
title: "Fitting extremes"
author: "Beni Stocker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting extremes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(extRemes)
library(here)
library(ggplot2)
library(patchwork)
```

## The data

Example data for four pixels
```{r}
load(here("data/df_cwd.RData"))
```

Create pixel ID and nest data by pixel.
```{r}
df <- df_cwd |>
  mutate(
    site = paste0("lon", as.character(lon), "_lat", as.character(lat))
  ) |>
  group_by(site) |>
  nest()
```

Visualise CWD maxima for two pixels.
```{r}
# pixel 1 is not extreme in 2023
df$data[[1]] |>
  ggplot(aes(year, max_deficit)) +
  geom_line()

# pixel 2 is extreme in 2023
df$data[[2]] |>
  ggplot(aes(year, max_deficit)) +
  geom_line()
```

## Fit extreme value distributions

Define some useful functions.
```{r}
# function to fit extreme value distribution
get_evd <- function(df, type = "GEV"){
  vals <- df |>
    pull(max_deficit)

  extRemes::fevd(
    x = vals,
    type = type,
    method = "MLE",
    units = "years"
    )
}

return_period <- c(2, 5, 10, 20, 30, 40, 50, 60, 70, 80, 90, 100)

# function to calculate return periods of each year in observations
get_return_periods <- function(df, mod){

  vals <- df |>
    pull(max_deficit)

  calc_return_period(
    vals,
    extract_loc(mod),
    extract_scale(mod)
    )
}

calc_return_period <- function(x, loc, scale){
  1 / (1 - exp(-exp(-(x-loc)/scale)))
}

extract_loc <- function(mod){
  loc <- mod$results$par[ "location" ]
  if (!is.null(loc)){
    return(loc)
  } else {
    return(NA)
  }
}

extract_scale <- function(mod){
  scale <- mod$results$par[ "scale" ]
  if (!is.null(scale)){
    return(scale)
  } else {
    return(NA)
  }
}
```

Fit.
```{r}
df <- df |>

  # # example for one pixel
  # ungroup() |>
  # slice(1) |>

  mutate(
    # fit GEV - pixel 2 causes a warning
    gev = purrr::map(data, ~get_evd(., type = "GEV")),

    # fit Gumbel - no warnings
    gumbel = purrr::map(data, ~get_evd(., type = "Gumbel"))
    ) |>

  # get return magnitudes for pre-defined set of return intervals
  mutate(
    
    return_magnitudes_gev    = purrr::map(
      gev,    
      ~unname(c(extRemes::return.level(., return.period = return_period)))),
    
    return_magnitudes_gumbel = purrr::map(
      gumbel, 
      ~unname(c(extRemes::return.level(., return.period = return_period))))
  ) |>

  # get return periods for all observations
  mutate(
    return_periods_gev    = purrr::map2(data, gev,    ~get_return_periods(.x, .y)),
    return_periods_gumbel = purrr::map2(data, gumbel, ~get_return_periods(.x, .y))
  ) |>

  # re-organise data: drop model objects and add return periods as columns to data
  select(-gev, -gumbel, -starts_with("return_magnitudes_")) |>
  unnest(c(data, return_periods_gev, return_periods_gumbel)) |>
  group_by(site) |>
  nest()
```

## Plot

Plots (bottom panels) show the expected return period for each year's observed CWD. 'Expected' is based on the fitted model. Models are fitted with all years' data.

### Pixel 1

Works ok - 2023 is not extreme
```{r}
gg1 <- df |>
  ungroup() |>
  slice(1) |>
  unnest(cols = c(data)) |>
  ggplot(aes(year, max_deficit)) +
  geom_line() +
  labs(x = "Year", y = "CWD (mm)") +
  theme_classic()

gg2 <- df |>
  ungroup() |>
  slice(1) |>
  unnest(cols = c(data)) |>
  ggplot(aes(year)) +
  geom_line(aes(y = return_periods_gev, color = "GEV")) +
  geom_line(aes(y = return_periods_gumbel, color = "Gumbel")) +
  labs(x = "Year", y = "Return period (yr)") +
  theme_classic()

gg1 / gg2
```

### Pixel 2

GEV fitting fails
```{r}
gg1 <- df |>
  ungroup() |>
  slice(2) |>
  unnest(cols = c(data)) |>
  ggplot(aes(year, max_deficit)) +
  geom_line() +
  labs(x = "Year", y = "CWD (mm)") +
  theme_classic()

gg2 <- df |>
  ungroup() |>
  slice(2) |>
  unnest(cols = c(data)) |>
  ggplot(aes(year)) +
  geom_line(aes(y = return_periods_gev, color = "GEV")) +
  geom_line(aes(y = return_periods_gumbel, color = "Gumbel")) +
  labs(x = "Year", y = "Return period (yr)") +
  theme_classic()

gg3 <- df |>
  ungroup() |>
  slice(2) |>
  unnest(cols = c(data)) |>
  ggplot(aes(year)) +
  geom_line(aes(y = return_periods_gumbel, color = "Gumbel")) +
  labs(x = "Year", y = "Return period (yr)") +
  theme_classic()

gg1 / gg2 / gg3
```

### Pixel 3

Works ok - 2023 is extreme.
```{r}
gg1 <- df |>
  ungroup() |>
  slice(3) |>
  unnest(cols = c(data)) |>
  ggplot(aes(year, max_deficit)) +
  geom_line() +
  labs(x = "Year", y = "CWD (mm)") +
  theme_classic()

gg2 <- df |>
  ungroup() |>
  slice(3) |>
  unnest(cols = c(data)) |>
  ggplot(aes(year)) +
  geom_line(aes(y = return_periods_gev, color = "GEV")) +
  geom_line(aes(y = return_periods_gumbel, color = "Gumbel")) +
  labs(x = "Year", y = "Return period (yr)") +
  theme_classic()

gg1 / gg2
```

### Pixel 4

GEV fails. 2023 is extreme.
```{r}
gg1 <- df |>
  ungroup() |>
  slice(4) |>
  unnest(cols = c(data)) |>
  ggplot(aes(year, max_deficit)) +
  geom_line() +
  labs(x = "Year", y = "CWD (mm)") +
  theme_classic()

gg2 <- df |>
  ungroup() |>
  slice(4) |>
  unnest(cols = c(data)) |>
  ggplot(aes(year)) +
  geom_line(aes(y = return_periods_gev, color = "GEV")) +
  geom_line(aes(y = return_periods_gumbel, color = "Gumbel")) +
  labs(x = "Year", y = "Return period (yr)") +
  theme_classic()

gg1 / gg2
```
